Molecular transport in human pial perivasculature

### 2024-11-12 paper/supplementary_main.tex
Clean-ups
Former-commit-id: 386f212cee0a5a09e05ddfa15b8e59d054bcc2c3 [formerly 3aaecf5b8769e98b3e4aa00676e63a48fee3ba03]
Former-commit-id: 7152b8a813ccaba34a88f72b36c2592d6eed8397

\subsection{Mathematical model definitions and further model specifications}
\label{sec:more_maths}

In these sections, we provide a complete definition of the
mathematical models considered.

\subsubsection{Notation and definitions}

@@

@@

\subsection{Numerical Method} \label{sec:details_numerical_method}

@@

\subsubsection{Solute transport} For the 3D concentration, we use broken polynomial spaces and interior penalty discontinuous Galerkin (DG) methods with upwinding in order to capture the sharp boundary layers.  For DG, we utilize the discrete space 
\begin{equation}
V_h^k = \{v_h \in L^2(\Omega): \;  v_h \vert_E \in \mathbb{P}^k(E), \; E \in \mathcal{T}_h\}.
\end{equation}
Recall that the 3D concentration $c$ is defined locally in each subdomain of $\Omega$ and satisfies
\begin{equation}
     \partial_t (\phi c) - \nabla \cdot (D \nabla (\phi c) ) + \nabla \cdot (\bm u c ) + \xi (\overline{c} - \hat c ) \delta_\Gamma = f \quad \quad \mathrm{in} \quad \Omega_{\rm CSF} \cup \Omega_{\rm{PAR}}.  \label{eq:3d_pde}
\end{equation}
The above is supplemented with the following boundary conditions. 
\begin{alignat}{2}
(D \nabla (\phi c) - \bm u c ) \cdot \bm{n}_{\rm SAS} &= g_{\mathrm{influx}} &&  \quad  \mathrm{on} \quad \spinal  \\ 
% (-D \nabla (\phi c) + \bm u c ) \cdot \bm{n}_{\spinal} &= 0 &&  \quad  \mathrm{on} \quad \spinal^{\mathrm{out}}  \\ 
(-D \nabla (\phi c) + \bm u c ) \cdot \bm{n}_{\rm skull} & =   \beta c  &&  \quad  \mathrm{on} \quad \Gamma_{\rm skull^{u}}  \\ 
(-D \nabla (\phi c) + \bm u c ) \cdot \bm{n}_{\rm skull} & =   0  &&  \quad  \mathrm{on} \quad \Gamma_{\rm skull^{l}}  \\ 
(-D \nabla (\phi c_{\rm{PAR}}) + \bm u_{\rm{PAR}} c_{\rm{PAR} }) \cdot \bm{n}_{\rm{Pia}} &= - (-D \nabla  c_{\rm{CSF}} + \bm u_{\rm CSF} c_{\rm CSF} ) \cdot \bm{n}_{\rm{Pia}} &&  \quad  \mathrm{on} \quad \pia \cup \Gamma_{\mathrm{LV}}  \\ 
(-D \nabla (\phi c_{\rm{PAR}}) + \bm u_{\rm{PAR}} c_{\rm{PAR} }) \cdot \bm{n}_{\rm{Pia}} & = \beta_{\rm{pia}} (c_{\rm PAR} - c_{\mathrm{CSF}}) &&  \quad  \mathrm{on} \quad \pia \cup \Gamma_{\mathrm{LV}}   .  
\end{alignat}

In the above, note that $\bm u \cdot \bm n  \neq 0$ on $\Gamma_{\rm LV} \cup \Gamma_{\rm skull^u}$ and $\bm u \cdot \bm n = 0$ on $\Gamma_{\rm SSAS} \cup \Gamma_{\rm skull^l} \cup \Gamma_{\rm Pia}$. To discretize the diffusion term, we utilize symmetric weighted interior penalty dG bilinear forms. We refer to \cite{ern2009discontinuous} and \cite[Section 4.5.2.3]{di2011mathematical} for details on this method.  We define 
\begin{multline}
a_h (c_h, v_h) = \sum_{E \in \mathcal{T}_h } \int_{E} D \nabla (\phi c_h) \cdot \nabla v_h + \sum_{F \in \mathcal{F}_i} \eta \frac{\gamma_{D,F}}{h_F} \int_{F} [c_h][v_h] 
- \sum_{F \in \mathcal{F}_i} \int_{F} \left( \{D \phi \nabla c_h \}_w \cdot \bm n_F [v_h]  + \{D \phi\nabla v_h \}_w \cdot \bm n_F [c_h] \right).
 \nonumber
 \end{multline}
In the above, for $F\in \mathcal{F}_h^i$ with a normal $n_F$ pointing from $E_F^1$ to $E_F^2$,  the jump is given by $[v_h] = v_h{}_{\vert_{E_F^1}} - v_h{}_{\vert_{E_{F}^2}}$, and the weighed average is defined as follows.
\begin{equation}
\{v_h\}_{w} = \frac{ \kappa_{\vert_{E_F^2}}}{\kappa_{\vert_{E_F^1}}+ \kappa_{\vert_{E_F^2}}} v_h{}_{\vert_{E_{F}^1}} + \frac{\kappa_{\vert_{E_F^1}}}{\kappa_{ \vert_{E_F^1}}+ \kappa_{\vert_{E_F^2}}} v_h{}_{\vert_{E_{F}^2}}, \quad \text{where} \quad \kappa = D \phi.  
\end{equation}
The parameter $\gamma_{D,F}$ is the harmonic mean of the diffusion coefficient given by 
$$
\gamma_{D,F} = \frac{ \kappa_{\vert_{E_F^1}}\kappa_{\vert_{E_F^2}}}{\kappa_{\vert_{E_F^1}}+ \kappa_{\vert_{E_F^2}}},
$$
and $\eta$ is a user specified penalty to be determined later.

>>>>

These sections provide more detailed descriptions of the mathematical
models and numerical approximations considered.

\subsection{Mathematical models and definitions}
\label{sec:more_maths}

\subsubsection{Notation}

In terms of domains, interfaces and boundaries, we consider the
parenchyma $\Omega_{\rm PAR} \subset \R^3$ and CSF spaces $\Omega_{\rm
  CSF} \subset \R^3$ with $\Omega = \Omega_{\rm PAR} \cup \Omega_{\rm
  CSF}$. The interface between the parenchyma and CSF spaces is
decomposed

\subsection*{Notation and left-overs}

\paragraph*{To be moved: notation and labels}
The thin pia membrane encloses the brain, and we hence denote the CSF-brain interface with $\Gamma_{\rm pia}$, except for the separately marked surface of the lateral ventricles $\Gamma_{\rm LV}$ (thus $\partial \Omega_{\rm CSF} \bigcap \partial \Omega_{\rm PAR} = \Gamma_{\rm LV} \cup \Gamma_{\rm Pia}$).

We denote the different parts of the outer boundary of $\Omega_{\rm
  CSF}$ with $\Gamma_{\mathrm{skull}}$ facing the dura and skull and
$\Gamma_{\rm SSAS}$ representing the lower interface towards the
spinal subarachnoid space (SSAS). $\Gamma_{\mathrm{skull}}$ is further
subdivided into its lower and upper parts: $\Gamma_{\mathrm{skull^l}}$
and $\Gamma_{\mathrm{skull^u}}$ (compare \Cref{fig:concept}). The
boundary of the spinal cord is denoted by $\Gamma_{\rm{SC}}$, which is
given by $\Gamma_{\rm{SC}} = \partial \Omega_{\rm PAR} \setminus
\partial \Omega_{\rm CSF}$ .

%% We denote the domain defined by the connected centerlines by
%% $\Lambda$.The vascular domain is then represented as the union of
%% cylindrical vessels of radius $R_1^i$ surrounding the centerlines
%% $\Lambda^i$. Moreover, we consider the surrounding perivascular spaces
%% as the union of annular cylinders of inner radius $R_1^i$ and outer
%% radius $R_2^i = \beta R_1^i$ of width $R_2^i - R_1^i = (\beta - 1)
%% R_1^i$.

@@

Recall that the 3D concentration $c$ is defined locally in each subdomain of $\Omega$ and satisfies
\begin{equation}
     \partial_t (\phi c) - \nabla \cdot (D \nabla (\phi c) ) + \nabla \cdot (\bm u c ) + \xi (\overline{c} - \hat c ) \delta_\Gamma = f \quad \quad \mathrm{in} \quad \Omega_{\rm CSF} \cup \Omega_{\rm{PAR}}.  \label{eq:3d_pde}
\end{equation}
The above is supplemented with the following boundary conditions. 
\begin{alignat}{2}
(D \nabla (\phi c) - \bm u c ) \cdot \bm{n}_{\rm SAS} &= g_{\mathrm{influx}} &&  \quad  \mathrm{on} \quad \spinal  \\ 
% (-D \nabla (\phi c) + \bm u c ) \cdot \bm{n}_{\spinal} &= 0 &&  \quad  \mathrm{on} \quad \spinal^{\mathrm{out}}  \\ 
(-D \nabla (\phi c) + \bm u c ) \cdot \bm{n}_{\rm skull} & =   \beta c  &&  \quad  \mathrm{on} \quad \Gamma_{\rm skull^{u}}  \\ 
(-D \nabla (\phi c) + \bm u c ) \cdot \bm{n}_{\rm skull} & =   0  &&  \quad  \mathrm{on} \quad \Gamma_{\rm skull^{l}}  \\ 
(-D \nabla (\phi c_{\rm{PAR}}) + \bm u_{\rm{PAR}} c_{\rm{PAR} }) \cdot \bm{n}_{\rm{Pia}} &= - (-D \nabla  c_{\rm{CSF}} + \bm u_{\rm CSF} c_{\rm CSF} ) \cdot \bm{n}_{\rm{Pia}} &&  \quad  \mathrm{on} \quad \pia \cup \Gamma_{\mathrm{LV}}  \\ 
(-D \nabla (\phi c_{\rm{PAR}}) + \bm u_{\rm{PAR}} c_{\rm{PAR} }) \cdot \bm{n}_{\rm{Pia}} & = \beta_{\rm{pia}} (c_{\rm PAR} - c_{\mathrm{CSF}}) &&  \quad  \mathrm{on} \quad \pia \cup \Gamma_{\mathrm{LV}}   .  
\end{alignat}

In the above, note that $\bm u \cdot \bm n  \neq 0$ on $\Gamma_{\rm LV} \cup \Gamma_{\rm skull^u}$ and $\bm u \cdot \bm n = 0$ on $\Gamma_{\rm SSAS} \cup \Gamma_{\rm skull^l} \cup \Gamma_{\rm Pia}$.

@@

\subsection{Numerical Method}
\label{sec:details_numerical_method}

@@

\subsubsection{Solute transport}
For the 3D concentration, we use broken polynomial spaces and interior penalty discontinuous Galerkin (DG) methods with upwinding in order to capture the sharp boundary layers.  For DG, we utilize the discrete space 
\begin{equation}
V_h^k = \{v_h \in L^2(\Omega): \;  v_h \vert_E \in \mathbb{P}^k(E), \; E \in \mathcal{T}_h\}.
\end{equation}
To discretize the diffusion term, we utilize symmetric weighted interior penalty dG bilinear forms. We refer to \cite{ern2009discontinuous} and \cite[Section 4.5.2.3]{di2011mathematical} for details on this method.  We define 
\begin{multline}
a_h (c_h, v_h) = \sum_{E \in \mathcal{T}_h } \int_{E} D \nabla (\phi c_h) \cdot \nabla v_h + \sum_{F \in \mathcal{F}_i} \eta \frac{\gamma_{D,F}}{h_F} \int_{F} [c_h][v_h] 
- \sum_{F \in \mathcal{F}_i} \int_{F} \left( \{D \phi \nabla c_h \}_w \cdot \bm n_F [v_h]  + \{D \phi\nabla v_h \}_w \cdot \bm n_F [c_h] \right).
 \nonumber
 \end{multline}
In the above, for $F\in \mathcal{F}_h^i$ with a normal $n_F$ pointing from $E_F^1$ to $E_F^2$,  the jump is given by $[v_h] = v_h{}_{\vert_{E_F^1}} - v_h{}_{\vert_{E_{F}^2}}$, and the weighed average is defined as follows.
\begin{equation}
\{v_h\}_{w} = \frac{ \kappa_{\vert_{E_F^2}}}{\kappa_{\vert_{E_F^1}}+ \kappa_{\vert_{E_F^2}}} v_h{}_{\vert_{E_{F}^1}} + \frac{\kappa_{\vert_{E_F^1}}}{\kappa_{ \vert_{E_F^1}}+ \kappa_{\vert_{E_F^2}}} v_h{}_{\vert_{E_{F}^2}}, \quad \text{where} \quad \kappa = D \phi.  
\end{equation}
The parameter $\gamma_{D,F}$ is the harmonic mean of the diffusion coefficient given by 
$$
\gamma_{D,F} = \frac{ \kappa_{\vert_{E_F^1}}\kappa_{\vert_{E_F^2}}}{\kappa_{\vert_{E_F^1}}+ \kappa_{\vert_{E_F^2}}},
$$
and $\eta$ is a user specified penalty to be determined later.
